Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-16 manuscript/climate_sensitivity_draft.Rmd
worked on height & CP fixes
#106 #111

We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$TWI$, crown position, and species traits as specified above.
We identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 
When a variable appeared in all of these models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction (Table 1). 
If the variable appeared in some but not all of these models, and its sign we was consistent across models, we considered this partial support/rejection. 
In presentation of the results below, we note instances where the $Rt_{ARIMA}$ model disagreed with the $Rt$ model, but otherwise do not discuss the $Rt_{ARIMA}$ model.

@@

Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually (Table 1; Fig. 4;). 
Specifically, $ln[H]$ appeared, with negative coefficient, in the best models and all models within $\Delta$AICc=2 of these (Tables S6-S7).
For the 1977 and 1999 droughts, $ln[H]$ did not appear in the best models, but was included, with negative coefficient, among some of the top models, which were statistically indistinguishable ($\Delta$AICc<2) from the top model (Tables 1, S6-S7).

Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). 
Crown position was never statistically significant (Table 1). 
It did, however, come out in the best models for 1977 and 1999 (Fig. 4), and was included in some of the top models for 1966 and all droughts combined (Table S6).
When $Rt_{ARIMA}$ was used as the predictor variable, crown position was never in a best model, and was included among the  top models only for 1977.
Crown position did not have a consistent influence, when included in top models, across years: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in all other years suppressed trees had the lowest $Rt$ (Fig. 4; Tables S6-S7).

@@

$Rt$ had a significantly negative response to $ln[TWI]$ in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in some of the models in 1999 and the $Rt_{ARIMA}$ model for 1966 (Fig. 4, Tables 4-5).
This negates the idea that trees in moist microsites would be less affected by drought. 
Nevertheless, we tested for a $ln[H] *ln[TWI]$ interaction, a negative sign of which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table. 
This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models  in which it appeared (Tables 1, S6-S7).
This term did appear, with positive coefficient, in the best $Rt_{ARIMA}$ model for all years combined (Table S7).

@@

Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. 4). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees. The negative effect of height on $Rt$ held after accounting for species' traits. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought resistance was not linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$). This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The directions of these responses were consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. 4; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig. **1a**, @helcoski_growing_2019) and had substantial but not devastating impacts on tree growth (Figs. **1b, 2**). These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S2**), severely stressed a larger portion of trees (Fig. **1b**). The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. **1b**), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-26% for each individual drought (Table S6).

Our analysis indicates that tree height has a stronger influence on drought response than does crown position (Fig. 4; Tables 1, S6-S7). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. However, the collinearity between the two variables (Fig. **3d**) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. **3d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. **3a-b**). Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves (*Campbell & Norman REF*). However, some decoupling between height and crown position is introduced by the configuration of neighboring trees (Fig. **3d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Fig. 4; Tables 1, S6-S7). Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].

Our analysis has the limitation that crown positions were recorded in 2018, as opposed to the years of the droughts. However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing crown positions would bias against the acceptance of our hypothesis. The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019]. It is safe to assume that currently suppressed trees were suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition [@sohn_potential_2016]. The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

>>>>

We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$TWI$ or crown position and $TWI$, and species traits as specified above.
Crown position and $ln[H]$ were never considered in the same model because of high correlation (0.73) between the two variables. 
We identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 
When a variable appeared in all of these models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction (Table 1). 
If the variable appeared in some but not all of these models, and its sign we was consistent across models, we considered this partial support/rejection. 
In presentation of the results below, we note instances where the $Rt_{ARIMA}$ model disagreed with the $Rt$ model, but otherwise do not discuss the $Rt_{ARIMA}$ model.

@@

Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually (Table 1; Fig. **4**). 
Specifically, $ln[H]$ appeared, with negative coefficient, in the best models and all top models (Tables S6-S7).
For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S7). 
For the 1999 drought, $ln[H]$ appeared in no top models for $Rt$ and some top models, with positive coefficient, for $Rt_{ARIMA}$.

Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). 
Crown position was not included in any of the top models for all droughts or for 1966 individually (Table 1). 
It did, however, come out in the best models for 1977 and 1999 (Fig. **4**; Table S6).
When $Rt_{ARIMA}$ was used as the response variable, crown position was never in a best model, and was included among the top models only for 1977 (Table S7).
Crown position did not have a consistent influence: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in 1999 intermediate and suppressed trees had lower $Rt$ (Fig. 4; Tables S6-S7).

@@

$Rt$ had a significantly negative response to $ln[TWI]$ in all drought years combined and in 1977 and 1999 individually (Fig. 4, Table S6). 
When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in some of the models in 1966 and 1999 (Table S7).
This negates the idea that trees in moist microsites would be less affected by drought. 
Nevertheless, we tested for a $ln[H] *ln[TWI]$ interaction, a negative sign of which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table. 
This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models in which it appeared (Tables 1, S6-S7).
This term did appear, with positive coefficient, in the best $Rt_{ARIMA}$ model for all years combined (Table S7).

@@

Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. 4). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019]. The negative effect of height on $Rt$ held after accounting for species' traits. We found only a marginal effect of crown exposure with inconsistent direction across two individual droughts.  There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought resistance was not linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$). This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The directions of these responses were consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. 4; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig. **1a**, @helcoski_growing_2019) and had substantial but not devastating impacts on tree growth (Figs. **1b, 2**). These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S2**), severely stressed a larger portion of trees (Fig. **1b**). The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. **1b**), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Table S6).

Our analysis indicates that tree height has an overall stronger influence on drought response than does crown position (Table 1). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. However, the collinearity between the two variables (Fig. **3d**) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. **3d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. **3a-b**). Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves (*Campbell & Norman REF*). *Previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019], and ...* However, some decoupling between height and crown position is introduced by the configuration of neighboring trees (Fig. **3d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Fig. 4; Tables 1). Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].

*Our analysis has the limitation that crown positions were recorded in 2018, as opposed to the years of the droughts. However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing crown positions would bias against the acceptance of our hypothesis. The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019]. It is safe to assume that currently suppressed trees were suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition [@sohn_potential_2016]. The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure. *
